---
title: "Analytics assignment 3"
format: pdf
editor: visual
---

## Cleaning Data

```{r}
library(tidyverse)
library(corrr)
library(factoextra)
library(FactoMineR)
library(NbClust)
library(cluster)
library(mclust)

dat <- Airbnb_Excel_Lab_AAMB_Co
```

## Sort out any missing values

```{r}
str(dat)
dim(dat) # 16317 40; before cleaning

head(dat)

#removed repeated rows
dat <- distinct(dat) # no duplicates were found

# removed the first 7 rows, did not seem relevant to what we are tyring to explore with the data 
dat <- dat[, -c(1,2,3,4,5,6, 7)] 

#sorting out categorical variables 
dat$Neighbourhood_Cleansed <- as.factor(dat$Neighbourhood_Cleansed)
dat$`Room_Type_Proper & Trim` <- as.factor(dat$`Room_Type_Proper & Trim`)

# roomtypeproper $ trim is the same as property type & proper trim so remove the first 1 
dat <- dat[, -4]

#missing values 
indices_missing <- which(is.na(dat))
missing_dat <- dat[indices_missing,]
head(missing_dat)
# NA'S were found for all the values there so will just remove those observations from the dataset. It was 52 observations which is less than 1% of our data set which is fine.

#removed missing values 
dat <-na.omit(dat)
dim(dat) # 16288 32 

#make dummys for categorical variables
#table(dat$`Room_Type_Proper & Trim`)

dat <- dat %>% 
  mutate(
    entire_home = as.integer(dat$`Room_Type_Proper & Trim` == "Entire Home/Apt"),
    hotel_room = as.integer(dat$`Room_Type_Proper & Trim` == "Hotel Room"),
    pvt_room = as.integer(dat$`Room_Type_Proper & Trim` == "Private Room"),
    shared_room = as.integer(dat$`Room_Type_Proper & Trim` == "Shared Room")
  )

dat <- dat %>% 
  select(- `Room_Type_Proper & Trim`)


#remove non revenue listings. They may not be beneficicial to what we are trying to determine. 
dat <- dat %>% 
  filter(Revenue_365 != 0) 
dim(dat) # 15851 35

#New variables 
# overall score 
dat <- dat %>% 
  mutate(
    avg_score = round((Review_Scores_Accuracy + Review_Scores_Cleanliness + Review_Scores_Checkin + Review_Scores_Communication + Review_Scores_Location + Review_Scores_Value)/6, 2)
  )

#occupancy rate 
dat <- dat %>% 
  mutate(
    occupancy_rate = round((Booked_365/365)*100, 2)
)


#ward averages for 70 of the wards that are not missing values and have active listings
Ward_avg <- dat %>% 
  group_by(Neighbourhood_Cleansed) %>% 
  summarise_all(., mean)

#remove Neighbourhood_Cleansed, we already have the geo co-ordinates
dat <- dat %>% 
  select(-Neighbourhood_Cleansed)


summary(dat)

highly_corr <- dat %>% 
  correlate() %>% 
  stretch() %>% 
  filter(abs(r) > 0.85) %>% 
  arrange(desc(abs(r)))
  
highly_corr

#remove availabily anf booked since we have occupancy rate
dat <- dat %>% 
  select(- c(Booked_30, Booked_60, Booked_90, Booked_365, Availability_30, Availability_60, Availability_90, Availability_365))

#remove the scores individually because captured by avg and scores rating
dat <- dat %>% 
  select(- c(Review_Scores_Accuracy, Review_Scores_Cleanliness, Review_Scores_Communication, Review_Scores_Location, Review_Scores_Checkin, Review_Scores_Value))

#remove entire home so there is no multicollinearity
dat <- dat[,-17]

save(dat, Ward_avg, file = "cleaning.RData")
load("cleaning.RData")
```

## Constructing PCA & Clustering Variables

Going to do both on the PC's and the variables and then we can decide which is better

Variables for clustering

- `Price`, `Revenue_365`, `Bedrooms`, `Bathrooms`, and `Accommodates`

- `Reviews_per_Month`

- `avg_score`, `occupancy_score`,

- `entire_home` , `pvt_room,`, `shared_room` and `hotel_room`

- `Latitude` , `Longitude`

```{r}
#for clustering variables
# put all these variables into 1 dataframe
clustering_variables <- data.frame(dat$Price, dat$Revenue_365, dat$Bedrooms, dat$Bathrooms, dat$Accommodates, dat$Reviews_per_Month, dat$avg_score, dat$occupancy_rate, dat$shared_room, dat$pvt_room, dat$hotel_room, dat$Latitude, dat$Longitude)

dim(clustering_variables) # 15851 13

#renamed the column names so that they didnt have dat before them
colnames(clustering_variables) <- c("Price", "Revenue_365", "Bedrooms", "Bathrooms", "Accommodates", "Reviews_per_Month", "avg_score", "occupancy_score", "shared_room", "pvt_room","hotel_room", "Latitude", "Longitude")

#scaled the variables 
scaled_clustering_vars <- scale(clustering_variables)
```

```{r}
#PCA
pca_output <- PCA(clustering_variables,scale.unit = T,graph = FALSE, ncp = 10)
fviz_eig(pca_output)

#choose the first 6 eigen values
inerita_perc <- pca_output$eig[,3]

#loadings
pca_loadings <- pca_output$var$coord[,1:6]

#scores
pca_scores <- pca_output$ind$coord[,1:6]

save(scaled_clustering_vars, pca_output, pca_scores, inerita_perc, file = "cluster vars and PCA.RData")
load("cluster vars and PCA.RData")
```

## Hierarchical Clustering

##### distance matrix calculation

```{r}

dist_matrix <- dist(scaled_clustering_vars, method = "euclidean")
```

##### PC's score distance matrix calculation:

```{r}
pca_dist_matrix <- dist(pca_scores, method = "euclidean")

save(dist_matrix, pca_dist_matrix, file = "distance matrices.RData")
load("distance matrices.RData")
```

### Complete Linkage

##### For variables

```{r}

var_complete_euc <- hclust(dist_matrix, method = "complete")

```

##### For PC's

```{r}
pca_complete_euc <- hclust(pca_dist_matrix, method = "complete")

save(var_complete_euc, pca_complete_euc, file = "Complete linkage full.RData")
load("Complete linkage full.RData")
```

### Determining where to cut the dendrogram

Decided to use NbClust to find how many clusters for Complete Linkage, the silhouette method, elbow method and Gap statistic.

#### subsetting the data

Decided to use a sample of the data to do the choosing of k as the HC methods are computationally exhaustive

```{r}
#the sample is 1k obs for each
set.seed(321)
sample_dat <- scaled_clustering_vars[sample(1:nrow(scaled_clustering_vars), 1000, replace = F),]

sample_pca <- pca_scores[sample(1:nrow(pca_scores), 1000, replace = F),]
```

#### NbClust

```{r}
nbclust_pca_complete <- NbClust(data = sample_pca,
                         distance = "euclidean",
                         min.nc = 2,
                         max.nc = 10,
                         method = "complete")
save(nbclust_pca_complete, file = "nbclust_complete.RData")
load("nbclust_complete.RData")
```

consensus: k = 2

#### Elbow method: PCA and cluster vars

```{r}
within_variation_manual <- function(X, cluster) {
  
  clusters <- unique(cluster)
  results <- data.frame(
    cluster = clusters,
    n = NA,
    within_variation = NA
  )
  
  total_within <- 0
  
  for (k in clusters) {
    
    # Points in cluster k
    Xk <- X[cluster == k, , drop = FALSE]
    
    # Cluster centroid
    centroid <- colMeans(Xk)
    
    # Squared distances to centroid
    squared_distances <- rowSums((sweep(Xk, 2, centroid, "-"))^2)
    
    # Within variation for cluster k
    Wk <- sum(squared_distances)
    
    results[results$cluster == k, "n"] <- nrow(Xk)
    results[results$cluster == k, "within_variation"] <- Wk
    
    total_within <- total_within + Wk
  }
  
  list(
    by_cluster = results,
    total_within_variation = total_within
  )
}
```

##### PCA_scores

```{r}
# complete
WSS_complete_pca <- numeric(9)
for (i in 2:10){
  clusters <- cutree(pca_complete_euc, k = i)
  result <- within_variation_manual(pca_scores, clusters)
  WSS_complete_pca[i] <- result$total_within_variation
}

WSS_complete_pca <- data.frame(
  k = 2:10,
  WSS = WSS_complete_pca[2:10]
)

ggplot(WSS_complete_pca, aes(x = k, y = WSS)) +
  geom_line(color = "#2c3e50", size = 1) +
  geom_point(color = "#e74c3c", size = 3) +
  scale_x_continuous(breaks = 2:10) +
  labs(
    title = "Elbow Method: Complete Linkage (PCA)",
    subtitle = "Analysis of Cape Town Airbnb Clusters",
    x = "Number of Clusters (k)",
    y = "Total Within-cluster Variation"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

```

##### cluster vars

```{r}
# complete
WSS_complete_vars <- numeric(9)
for (i in 2:10){
  clusters <- cutree(pca_complete_euc, k = i)
  result <- within_variation_manual(scaled_clustering_vars, clusters)
  WSS_complete_vars[i] <- result$total_within_variation
}

WSS_complete_vars <- data.frame(
  k = 2:10,
  WSS = WSS_complete_vars[2:10]
)

ggplot(WSS_complete_vars, aes(x = k, y = WSS)) +
  geom_line(color = "#2c3e50", size = 1) +
  geom_point(color = "#e74c3c", size = 3) +
  scale_x_continuous(breaks = 2:10) +
  labs(
    title = "Elbow Method: Complete Linkage (Clustering Vars)",
    subtitle = "Analysis of Cape Town Airbnb Clusters",
    x = "Number of Clusters (k)",
    y = "Total Within-cluster Variation"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

```

consensus: k = 6

#### Sillohuete method

##### pca

```{r}
#complete
fviz_nbclust(sample_pca, hcut, method = "silhouette", hc_method = "complete")

```

##### cluster vars

```{r}
#complete
fviz_nbclust(sample_dat, hcut, method = "silhouette", hc_method = "complete")

```

consensus: k =2

#### Gap stat

##### pca

```{r}
set.seed(321)
#complete
gap_complete <- clusGap(
  sample_pca,
  FUNcluster = hcut,
  hc_method = "complete",
  K.max = 10,
  B = 50
)

fviz_gap_stat(gap_complete)

```

##### cluster vars

```{r}
set.seed(321)
#complete
gap_complete_vars <- clusGap(
  sample_dat,
  FUNcluster = hcut,
  hc_method = "complete",
  K.max = 10,
  B = 50
)

fviz_gap_stat(gap_complete_vars)

save(gap_complete, gap_complete_vars, file = "Gap stats complete.RData")
load("Gap stats complete.RData")
```

consensus: k = 2

## Cluster Validation Summary

The determination of the optimal number of clusters ($k$) involved a comparative analysis across three hierarchical linkage methods. The results below illustrate the mathematical sensitivity of the Cape Town Airbnb dataset to different clustering algorithms.

| Validation Method Complete Linkage            |       |
|-----------------------------------------------|:-----:|
| **NbClust (Majority Vote pca sample scores)** | $k=2$ |
| **Elbow Method (WSS)**                        | $k=6$ |
| **Average Silhouette**                        | $k=2$ |
| **Gap Statistic**                             | $k=1$ |

## Hierarchical Cluster Selection and Validation

Based on a comparative analysis of formal validation metrics, $k=2$ was selected as the optimal number of clusters for the hierarchical Complete Linkage model. While the Elbow method indicated finer structural nuances at $k=6$, a two-cluster solution provides the most statistically defensible and stable partition for the Cape Town Airbnb dataset.

### Rationale for $k=2$

The selection of $k=2$ is supported by the following statistical evidence:

- **Majority Validation:** Both the **NbClust** majority vote and the **Average Silhouette** analysis identified $k=2$ as the most stable and well-separated global partition for this dataset.
- **Statistical Parsimony:** Following the principle of Occam's Razor, $k=2$ is the most parsimonious model that explains the data, particularly as the **Gap Statistic** suggested a lack of complex global clustering ($k=1$).
- **Market Interpretation:** Given the large sample size of \~15.8k observations, $k=2$ effectively captures the primary macro-structure of the market—separating "Standard" listings from "Niche/Elite" properties—without over-segmenting based on minor variances.

### Comparative Perspective

The hierarchical approach provides a broad macro-view of the Cape Town landscape. While it favors a simplified $k=2$ structure based on global distance measures, this serves as a critical baseline for comparison against density-based methods, which are more sensitive to local market concentrations.

## Fitting complete Linkage on full pca_scores and creating 2 clusters and identifying algorithm

### Cutting the tree

##### PCA

```{r}

cut_pca_tree <- cutree(pca_complete_euc, k = 2)
fviz_cluster(list(data = pca_scores, cluster = cut_pca_tree),
             geom = "point", 
             ellipse.type = "convex", 
             ggtheme = theme_minimal(),
             main = "Cluster Visualization (Complete Linkage, k=2)")

```

##### Clustering vars

```{r}

cut_vars_tree <- cutree(var_complete_euc, k = 2)
fviz_cluster(list(data = scaled_clustering_vars, cluster = cut_vars_tree),
             geom = "point", 
             ellipse.type = "convex", 
             ggtheme = theme_minimal(),
             main = "Cluster Visualization (Complete Linkage, k=2)")
```

### sillhoutte method

##### PCA

```{r}
sil_obj_pca <- silhouette(cut_pca_tree, pca_dist_matrix)


fviz_silhouette(sil_obj_pca) +
  labs(title = "Silhouette Plot: Complete Linkage (k=2)")

summary(sil_obj_pca)

```

##### Clustering Vars

```{r}
sil_obj_vars <- silhouette(cut_vars_tree, dist_matrix)


fviz_silhouette(sil_obj_vars) +
  labs(title = "Silhouette Plot: Complete Linkage (k=2)")


summary(sil_obj_vars)
save(sil_obj_pca, sil_obj_vars, file = "sil complete.RData")
load("sil complete.RData")
```

### Cophenetic

##### PCA

```{r}

cophenetic_cors <- numeric(1)

coph_dist_com <- cophenetic(pca_complete_euc)
cophenetic_cors[1]<- cor(pca_dist_matrix, coph_dist_com)

names(cophenetic_cors) <- c("complete")

cophenetic_cors
```

##### Clustering Vars

```{r}
cophenetic_cors <- numeric(1)

coph_dist_com <- cophenetic(var_complete_euc)
cophenetic_cors[1]<- cor(dist_matrix, coph_dist_com)

names(cophenetic_cors) <- c("complete")

cophenetic_cors
```

### Profiling

#### PCA

```{r}
pca_complete_df <- data.frame(pca_scores, cluster = cut_pca_tree)

# Calculates the means for every numeric column
pca_complete_means <- aggregate(. ~ cluster, data = pca_complete_df, FUN = mean)

# View the table
print(pca_complete_means)
```

```{r}
plot_data <- pca_complete_means%>%
  mutate(Market_Segment = ifelse(cluster == 1, "Standard", "Elite")) %>%
  pivot_longer(cols = starts_with("Dim"), 
               names_to = "Dimension", 
               values_to = "Mean_Score")

# 2. Create the plot
ggplot(plot_data, aes(x = Dimension, y = Mean_Score, group = Market_Segment, color = Market_Segment)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.5) +
  scale_color_manual(values = c("Standard" = "blue", "Elite" = "red")) +
  labs(title = "Hierarchical Cluster Profiles: PCA Dimension Centroids",
       subtitle = "Complete Linkage (k=2) separation of the Cape Town Airbnb Market",
       x = "Principal Components",
       y = "Average Centroid Score",
       color = "Market Segment") +
  theme_minimal() +
  theme(legend.position = "top")
```

##### Cluster 1: The Standard Market (Baseline)

Representing the vast majority of the Cape Town dataset, this cluster defines the "average" listing. \* **Property Scale:** Standard-sized residential units (typically 1–3 bedrooms). \* **Commercial Profile:** Moderate occupancy and revenue, reflecting typical secondary-income or "mom-and-pop" rentals. \* **Positioning:** Centered at the global mean of the PCA space, these listings represent the high-density core of the market.

##### Cluster 2: The Ultra-Luxury Outliers (Elite Tier)

This segment represents a statistically extreme group of "super-outliers" defined by their distance from the market average. \* **Exceptional Scale (Dim.1):** With a score of +24.67, these are properties of massive proportions (e.g., 8+ bedroom villas and mansions) that carry extreme price premiums. \* **Commercial Intensity (Dim.2 & 3):** High positive scores suggest these are professionally managed, high-revenue assets, likely concentrated in prime geographic hotspots. \* **Expectation Gap (Dim.4):** Notably, Cluster 2 displays a negative score for Dimension 4. This indicates that despite their luxury status, these listings receive lower average ratings. This likely reflects a "critical guest" phenomenon, where guests paying extreme premiums hold significantly higher expectations than the market average.

#### Clustering Vars

```{r}
vars_complete_df <- data.frame(clustering_variables, cluster = cut_pca_tree)

# Calculates the means for every numeric column
vars_complete_means <- aggregate(. ~ cluster, data = vars_complete_df, FUN = mean)

# View the table
print(vars_complete_means)
```

##### Segment 1: The Standard Market (Majority Baseline)

Cluster 1 represents the vast majority of the dataset and serves as the baseline for the Cape Town rental market. \* **Property Scale:** These are typical residential offerings, averaging **1.8 bedrooms** and **1.68 bathrooms**, accommodating roughly **3.7 guests**. \* **Pricing & Revenue:** With an average nightly price of **R 2,238.89**, these listings generate an estimated annual revenue of **R 320,270.10**. \* **Performance:** This segment maintains a high average guest score of **4.79**, suggesting a reliable and satisfactory experience for the general traveler. \* **Geographic Positioning:** Located at an average longitude of **18.46**, these properties are distributed across the central and suburban areas of the city.

##### Segment 2: The Ultra-Luxury Commercial Tier (The Elite)

Cluster 2 identifies a small, statistically extreme segment of "super-properties" that operate on a completely different economic scale. \* **Exceptional Scale:** This tier consists of massive estates or boutique offerings averaging **14.25 bedrooms and bathrooms**. \* **Financial Dominance:** The pricing is extreme, averaging **R 57,041.00 per night**, which drives an average annual revenue of over **R 16.4 million**. \* **High-Intensity Management:** Despite the luxury price point, these properties maintain a **66% occupancy rate**, significantly higher than the 39% seen in the standard market. This suggests they are professional commercial assets rather than casual rentals. \* **The "Critical Guest" Phenomenon:** Notably, the average rating drops to **4.32**. This indicates that guests paying premium rates have exponentially higher expectations and are far more critical in their evaluations than those in the standard market. \* **Prime Location:** The shift in average longitude to **18.39** confirms a concentration in the high-value Atlantic Seaboard corridor.

```{r}
library(sf)
library(tmap)
library(dplyr)

map_data <- clustering_variables %>%
  select(Longitude, Latitude) %>% 
  mutate(Segment = factor(cut_vars_tree, 
                          levels = c(1, 2), 
                          labels = c("Standard", "Elite"))) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# 2. Reset tmap and plot
tmap_mode("plot")

tm_shape(map_data) +
  tm_dots(col = "Segment", 
          palette = c("blue", "red"),
          size = 0.09,
          alpha = 0.6,
          title = "Airbnb Segments") +
  tm_layout(title = "Cape Town Market Clusters",
            legend.outside = TRUE,
            frame = FALSE)

tmap_mode("view") 


tm_shape(map_data) +
  tm_dots(col = "Segment", 
          palette = c("Standard" = "blue", "Elite" = "pink"),
          size = 0.09, 
          alpha = 0.7,
          title = "Airbnb Segments") + 
  tm_view(set.view = c(18.4233, -33.9249, 11)) 
```
