---
title: "Analytics assignment 3"
format: pdf
editor: visual
---

## Cleaning Data

```{r}
library(tidyverse)
library(corrr)
dat <- Airbnb_Excel_Lab_AAMB_Co
```

## Sort out any missing values

```{r}
str(dat)
dim(dat) # 16317 40; before cleaning

head(dat)

#removed repeated rows
dat <- distinct(dat) # no duplicates were found

# removed the first 7 rows, did not seem relevant to what we are tyring to explore with the data 
dat <- dat[, -c(1,2,3,4,5,6, 7)] 

#sorting out categorical variables 
dat$Neighbourhood_Cleansed <- as.factor(dat$Neighbourhood_Cleansed)
dat$`Room_Type_Proper & Trim` <- as.factor(dat$`Room_Type_Proper & Trim`)

# roomtypeproper $ trim is the same as property type & proper trim so remove the first 1 
dat <- dat[, -4]

#missing values 
indices_missing <- which(is.na(dat))
missing_dat <- dat[indices_missing,]
head(missing_dat)
# NA'S were found for all the values there so will just remove those observations from the dataset. It was 52 observations which is less than 1% of our data set which is fine.

#removed missing values 
dat <-na.omit(dat)
dim(dat) # 16288 32 

#make dummys for categorical variables
#table(dat$`Room_Type_Proper & Trim`)

dat <- dat %>% 
  mutate(
    entire_home = as.integer(dat$`Room_Type_Proper & Trim` == "Entire Home/Apt"),
    hotel_room = as.integer(dat$`Room_Type_Proper & Trim` == "Hotel Room"),
    pvt_room = as.integer(dat$`Room_Type_Proper & Trim` == "Private Room"),
    shared_room = as.integer(dat$`Room_Type_Proper & Trim` == "Shared Room")
  )

dat <- dat %>% 
  select(- `Room_Type_Proper & Trim`)


#remove non revenue listings. They may not be beneficicial to what we are trying to determine. 
dat <- dat %>% 
  filter(Revenue_365 != 0) 
dim(dat) # 15851 35

#New variables 
# overall score 
dat <- dat %>% 
  mutate(
    avg_score = round((Review_Scores_Accuracy + Review_Scores_Cleanliness + Review_Scores_Checkin + Review_Scores_Communication + Review_Scores_Location + Review_Scores_Value)/6, 2)
  )

#occupancy rate 
dat <- dat %>% 
  mutate(
    occupancy_rate = round((Booked_365/365)*100, 2)
)


#ward averages for 70 of the wards that are not missing values and have active listings
Ward_avg <- dat %>% 
  group_by(Neighbourhood_Cleansed) %>% 
  summarise_all(., mean)

#remove Neighbourhood_Cleansed, we already have the geo co-ordinates
dat <- dat %>% 
  select(-Neighbourhood_Cleansed)


summary(dat)

highly_corr <- dat %>% 
  correlate() %>% 
  stretch() %>% 
  filter(abs(r) > 0.85) %>% 
  arrange(desc(abs(r)))
  
highly_corr

#remove availabily anf booked since we have occupancy rate
dat <- dat %>% 
  select(- c(Booked_30, Booked_60, Booked_90, Booked_365, Availability_30, Availability_60, Availability_90, Availability_365))

#remove the scores individually because captured by avg and scores rating
dat <- dat %>% 
  select(- c(Review_Scores_Accuracy, Review_Scores_Cleanliness, Review_Scores_Communication, Review_Scores_Location, Review_Scores_Checkin, Review_Scores_Value))


```

## PCA 

```{r}

pca_output <- prcomp(dat, center = TRUE, scale. = TRUE)
summary(pca_output)

plot(pca_output$sdev, type = "b")

vars <- pca_output$sdev^2
prop_var <- vars / sum(vars)
cum_var <- cumsum(prop_var) # choose the first 2 PC's

loadings <- pca_output$rotation[,1:6]


```

Here is your **Component Breakdown**:

### **PC1: The "Scale & Revenue" Axis (31-35% Loading)**

This is your "Big Business" component.

- **Top Loadings:** `Price`, `Revenue_365`, `Bedrooms`, `Bathrooms`, and `Accommodates`.

- **Interpretation:** This component separates high-value, large properties from smaller, cheaper ones. Since it is heavily loaded on both size and revenue, it represents the **Economic Power** of the listing.

### **PC2: The "Popularity & Activity" Axis (40-50% Loading)**

This is the "Hustle" component.

- **Top Loadings:** `Number_of_reviews`, `Number_of_Reviews_ltm`, and `Reviews_per_Month`.

- **Interpretation:** Interestingly, these loadings are **negative** (-0.40 to -0.50). This just means that as you move *left* on the PC2 axis, the listings become much more "active" with higher turnover. It captures how frequently a place is being booked, regardless of its price.

### **PC3: The "Short-Term Yield" Axis**

- **Top Loadings:** `Revenue_30`, `Revenue_60`, and `Revenue_90`.

- **Interpretation:** Notice how these are grouped differently than `Revenue_365`. This PC captures listings that are currently "hot" or performing very well in the short term, versus those that have steady long-term annual income.

### **PC4: The "Guest Experience" Axis**

- **Top Loadings:** `Review_Scores_Rating` (-0.63) and `avg_score` (-0.63).

- **Interpretation:** This is almost purely about **Quality**. It separates the "Superhosts" from the lower-rated listings.

### **PC5: The "Property Type" Axis**

- **Top Loadings:** `entire_home` (0.52) vs. `pvt_room` (-0.50).

- **Interpretation:** This is a "Binary Toggle." It distinguishes between listings that offer the whole house versus those offering just a room.

### **PC6: The "Spatial" Axis**

- **Top Loadings:** `Latitude` (-0.62) and `Longitude` (0.69).

- **Interpretation:** This is your **Geography** driver. In Cape Town, this likely represents the "Atlantic Seaboard vs. The Rest" or "City vs. Suburbs" split.

**recommend using PC1 through PC5.**

- **PC1:** How big/expensive?

- **PC2:** How busy?

- **PC3:** Current demand?

- **PC4:** How good is the service?

- **PC5:** House or Room?
