---
title: "Analytics Assignment 3 - Cluster Analysis"
author: "Natalie Chidikwindi & Lihle Sigenu"
format: pdf
editor: visual
---

```{r}
#| label: libraries & read data
#| warning: false
#| include: false

# install.packages("dplyr")
library(dplyr)
# install.packages("ggcorrplot")
library(ggcorrplot)
library(ggplot)

listings <- Airbnb_Excel_Lab_AAMB_Co
```

# Exploratory Data Analysis

Any analysis starts with knowing your data, therefore in this part of the assignment, you are expected to check all your variables first, use str() to see the types of your variables. You are also expected to detect any discrepancies/anomalies/missing values in your dataset using any or all of the following: tables, summaries, frequencies, plots. If need be, you may apply some dimension reduction techniques.

Some useful data preprocessing:

-   clean and variable mutate

-   dummy variables

-   data overview

-   ward level averages

```{r}
#| label: na-checking
#| echo: false
#| warning: false

################################################################
# Create a table containing variable type and % of missing values
################################################################
vars <- colnames(listings) # variable names

# Create variable of # of missing variables in each column
missing <- c()
var_type <- c()
for (i in 1:(dim(listings)[2])) {
  missing[i] <- sum(is.na(listings[,i]))
  var_type <- class(listings[,i])
}
missing_table <- data.frame(variable = vars, missing_vals = missing, type = var_type)

# Drop Neighbourhood proper and trim
listings <- listings %>% 
  select(-`Neighbourhood_Proper & Trim`)

# Drop na obs
listings <- na.omit(listings)
```

```{r}
#| label: variable summaries
#| echo: false
#| warning: false

# Create a table of data summaries
sum_matrix <- matrix(NA, nrow = dim(listings)[2] - 12, ncol = 4)
colnames(sum_matrix) <- c("Min", "Median", "Mean", "Max")
rownames(sum_matrix) <- names(listings)[13:dim(listings)[2]]

for (i in 13:dim(listings)[2]) {

  x <- unlist(listings[, i])

  sum_matrix[i - 12, ] <- c(
    min(x, na.rm = TRUE),
    median(x, na.rm = TRUE),
    round(mean(x, na.rm = TRUE),3),
    max(x, na.rm = TRUE)
  )
}
skewness <- sum_matrix[,3] - sum_matrix[,2]
sum_matrix <- cbind(sum_matrix, skewness)
# Problem is that private rooms are listed as having 50 beds/baths
# Max bathrooms is 50 because all bathrooms in a B&B are listed
# Max bedrooms is 50 also as above
```

```{r}
#| label: plots
#| echo: false
#| warning: false

# Histogram of prices
hist(listings$Price)

# Latitude vs Longitude
plot(listings$Longitude ~ listings$Latitude)

# Checking average variables for different wards
ward_data <- listings %>%  
  group_by(Neighbourhood_Cleansed) %>% 
  summarise(AvgPrice = mean(Price, na.rm = T),
            AvgBedrooms = mean(Bedrooms, na.rm = T),
            AvgBathrooms = mean(Bathrooms, na.rm = T),
            AvgCapacity = mean(Accommodates, na.rm = T),
            AvgRating = mean(Review_Scores_Value, na.rm = T)) %>% 
  arrange(-AvgPrice)

# Plot the average price for wards against other variables
plot(ward_data$AvgPrice ~ ward_data$AvgBedrooms)
plot(ward_data$AvgPrice ~ ward_data$AvgBathrooms)
plot(ward_data$AvgPrice ~ ward_data$AvgCapacity)
plot(ward_data$AvgPrice ~ ward_data$AvgRating)
```

```{r}
#| label: corr-plots
#| echo: false
#| warning: false

# Corr plot between ward average vars
cor(ward_data[,-1])

# Corr plot between original vars
numeric_vars <- listings %>% 
  select(-c("ID","Picture_URL", "Host_ID", "Host_Name_Trimmed & Proper", "Neighbourhood_Cleansed", "Property_Type_Proper & Trim", "Room_Type_Proper & Trim"))
cor_numerics <- cor(numeric_vars)
ggcorrplot(cor_numerics, type = "lower") +
  theme(axis.text.x = element_text(angle = 90))
```
