---
title: "Analytics Assignment 3"
subtitle: "Cluster Analysis: An investigation into AirBnb Listings in Cape Town"
author: "Natalie Chidikwindi & Lihle Sigenu"
format: pdf
editor: visual
---

# Preliminary

```{r}
# Install libraries
library(tidyverse)
library(corrr)
library(kohonen)
library(cluster)
library(NbClust)
library(factoextra)
library(FactoMineR)
library(mclust)
library(fpc)
library(dbscan)
library(viridis)
library(sf)
library(tmap)
library(RColorBrewer)
library(gridExtra)
library(fossil)
# Load data 
dat <- Airbnb_Excel_Lab_AAMB_Co
```

# Exploratory Data Analysis

## Data Cleaning

```{r}
#| label: data-cleaning
#| warning: false
#| include: false

# Checking variable types and information
str(dat)
listings <- dat 

# Dimensionality 
dim(dat) # 16317 40; before cleaning
head(dat)

# Remove duplicates
dat <- distinct(dat) # no duplicates were found

dat <- dat %>%
  filter(!(Bedrooms > 0 & Beds == 0)) %>%  # Remove rooms with bedrooms but no beds
  filter(!(Bedrooms == 0 & Beds == 0))  # Remove obvious data errors

# Removed the first 7 rows, did not seem relevant to what we are tyring to explore with the data 
dat <- dat[, -c(1,2,3,4,5,6, 7)] 

# Sorting out categorical variables 
dat$Neighbourhood_Cleansed <- as.factor(dat$Neighbourhood_Cleansed)
dat$`Room_Type_Proper & Trim` <- as.factor(dat$`Room_Type_Proper & Trim`)

# roomtypeproper $ trim is the same as property type & proper trim so remove the first 1 
dat <- dat[, -4]

# Checking for missing values 
indices_missing <- which(is.na(dat))
missing_dat <- dat[indices_missing,]
head(missing_dat)
# NA'S were found for all the values there so will just remove those observations from the dataset. It was 52 observations which is less than 1% of our data set which is fine.

#removed missing values 
dat <-na.omit(dat)
dim(dat) # 15228 32 after removing missing values 

# Make dummies for categorical variables

dat <- dat %>% 
  mutate(
    entire_home = as.integer(dat$`Room_Type_Proper & Trim` == "Entire Home/Apt"),
    hotel_room = as.integer(dat$`Room_Type_Proper & Trim` == "Hotel Room"),
    pvt_room = as.integer(dat$`Room_Type_Proper & Trim` == "Private Room"),
    shared_room = as.integer(dat$`Room_Type_Proper & Trim` == "Shared Room")
  )

dat <- dat %>% 
  select(- `Room_Type_Proper & Trim`)


# Remove non revenue listings. They may not be beneficicial to what we are trying to determine. 
dat <- dat %>% 
  filter(Revenue_365 != 0) 
dim(dat) # 14806 35

#remove the 2 outliers where theres 50 beds and 50 baths and 32 and 10

dat <- dat %>%
  filter(
    Bedrooms < 10,                 
    Bathrooms < 10,                
  )
#####################
# Feature engineering
#####################

# overall review score 
dat <- dat %>% 
  mutate(
    avg_score = round((Review_Scores_Accuracy + Review_Scores_Cleanliness + Review_Scores_Checkin + Review_Scores_Communication + Review_Scores_Location + Review_Scores_Value)/6, 2)
  )

# Occupancy rate 
dat <- dat %>% 
  mutate(
    occupancy_rate = round((Booked_365/365)*100, 2)
)


#ward averages for 70 of the wards that are not missing values and have active listings
Ward_avg <- dat %>% 
  group_by(Neighbourhood_Cleansed) %>% 
  summarise_all(., mean)

#remove Neighbourhood_Cleansed, we already have the geo co-ordinates
dat <- dat %>% 
  select(-Neighbourhood_Cleansed)

# Checking highly correlated variables
highly_corr <- dat %>% 
  correlate() %>% 
  stretch() %>% 
  filter(abs(r) > 0.85) %>% 
  arrange(desc(abs(r)))

#remove availabily and booked since we have occupancy rate
dat <- dat %>% 
  select(- c(Booked_30, Booked_60, Booked_90, Booked_365, Availability_30, Availability_60, Availability_90, Availability_365))

#remove the scores individually because captured by avg and scores rating
dat <- dat %>% 
  select(- c(Review_Scores_Accuracy, Review_Scores_Cleanliness, Review_Scores_Communication, Review_Scores_Location, Review_Scores_Checkin, Review_Scores_Value))

#remove entire home so there is no multicollinearity
dat <- dat[,-17]
```

## Visualization

```{r}
#| label: fig-plots
#| echo: false
#| warning: false

# Histogram of prices
p0 <- ggplot(listings, aes(x = Price)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  labs(
    title = "Dist of Listing Prices",
    x = "Daily Price"
  ) +
  theme_minimal()


# Checking average variables for different wards
ward_data <- listings %>%  
  group_by(Neighbourhood_Cleansed) %>% 
  summarise(AvgPrice = mean(Price, na.rm = T),
            AvgBedrooms = mean(Bedrooms, na.rm = T),
            AvgBathrooms = mean(Bathrooms, na.rm = T),
            AvgCapacity = mean(Accommodates, na.rm = T),
            AvgRating = mean(Review_Scores_Value, na.rm = T),
            AvgLat = mean(Latitude, na.rm =T),
            AvgLong = mean(Longitude, na.rm = T),
            AvgBooked = mean(Booked_365, na.rm = T)) %>% 
  arrange(-AvgPrice)
# Plot the average price for wards against other variables
p1 <- ggplot(ward_data, aes(x = AvgCapacity, y = AvgPrice)) +
  geom_point(color = "#e74c3c", size = 1.5) +
  labs(
    title = "Capacity vs Ward Price",
    x = "Average Capacity",
    y = "Average Ward Price"
  ) +
  theme_minimal()

p2 <- ggplot(ward_data, aes(x = AvgRating, y = AvgPrice)) +
  geom_point(color = "blue", size = 1.5) +
  labs(
    title = "Rating vs Ward Price",
    x = "Average Rating",
    y = "Average Ward Price"
  ) +
  theme_minimal()

p3 <- ggplot(ward_data, aes(x = AvgBooked, y = AvgPrice)) +
  geom_point(color = "green", size = 1.5) +
  labs(
    title = "# Booked Days vs Ward Price",
    x = "Average Annual # of Booked Days",
    y = "Average Ward Price"
  ) +
  theme_minimal()

grid.arrange(
  p0, p1, p2, p3,
  ncol = 2
)
```

```{r}
#| label: fig-tmap
#| echo: false
#| warning: false

price_bins <- cut(round(ward_data$AvgPrice,2), breaks = 5)

#OR ON A TMAP
ward_sf <- st_as_sf(
  ward_data,
  coords = c("AvgLong", "AvgLat"),
  crs = 4326
)
tm_shape(ward_sf) +
  tm_dots(
    col = "AvgPrice",
    palette = "-RdBu",
    size = 0.5,
    style = "cont"
  ) +
  tm_layout(
    legend.outside = TRUE
  ) + 
  tm_basemap("Esri.WorldGrayCanvas")
```

## Principle Component Analysis

```{r}
#| label: pca
#| echo: true
#| warning: false

# Performing PCA on selected variables
pca_output <- PCA(dat,scale.unit = T,graph = FALSE, ncp = 10)
fviz_eig(pca_output, main = "Scree plot of Principle Components",
         geom = "line",
         xlab = "Number of Principle Components")

#choose the first 6 eigen values
inerita_perc <- pca_output$eig[,3]

#loadings
pca_loadings <- pca_output$var$coord[,1:5]

#scores
pca_scores <- pca_output$ind$coord[,1:5]
```

## Clustering Variables

For clustering purposes, the following variables have been selected:

-   `Price`, `Bedrooms`, `Bathrooms`

-   `Reviews_per_Month`

-   `avg_score`, `occupancy_score`,

-   `pvt_room,`, `shared_room` and `hotel_room`

```{r}
#| label: cluster-vars
#| echo: true

#for clustering variables
# put all these variables into 1 dataframe
clustering_variables <- data.frame(dat$Price, dat$Bedrooms, dat$Bathrooms, dat$Reviews_per_Month, dat$avg_score, dat$occupancy_rate, dat$shared_room, dat$pvt_room, dat$hotel_room)

# # Original variables considered
clustering_var_orig <- data.frame(dat$Price, dat$Bedrooms, dat$Bathrooms, dat$Reviews_per_Month, dat$avg_score, dat$occupancy_rate, dat$shared_room, dat$pvt_room, dat$hotel_room, dat$Latitude, dat$Longitude)
# dim(clustering_variables) # 14806 9

#renamed the column names so that they didnt have dat before them
colnames(clustering_variables) <- c("Price", "Bedrooms", "Bathrooms", "Reviews_per_Month", "avg_score", "occupancy_score", "shared_room", "pvt_room","hotel_room")

colnames(clustering_var_orig) <- c("Price", "Bedrooms", "Bathrooms", "Reviews_per_Month", "avg_score", "occupancy_score", "shared_room", "pvt_room","hotel_room", "Latitude", "Longitude")

# Scaled the variables 
scaled_clustering_vars <- scale(clustering_variables)
```

# Hierarchical Clustering

## Distance Matrix

```{r}
#| label: dist-matrix

# Compute distance matrix using eucl dist
dist_matrix <- dist(scaled_clustering_vars, method = "euclidean")
```

## Complete Linkage

```{r}
#| label: complete-linkage
#| echo: true
#| warning: false

# Computing complete linkage on distance matrix
var_complete_euc <- hclust(dist_matrix, method = "complete")
```

### Determining where to cut the dendrogram

Decided to use NbClust to find how many clusters for Complete Linkage, the silhouette method, elbow method and Gap statistic.

#### Subsetting the data

Decided to use a sample of the data to do the choosing of k as the HC methods are computationally exhaustive

```{r}
#| label: sub-sampling
#| echo: true

#the sample is 1k obs for each
set.seed(321)
clustering_variables <- clustering_variables %>%
  mutate(room_type = case_when(
    pvt_room == 1 ~ "private",
    shared_room == 1 ~ "shared",
    hotel_room == 1 ~ "hotel",
    TRUE ~ "other"
  ))

# Stratified sample - proportional to room types
sampled_data <- clustering_variables %>%
  group_by(room_type) %>%
  slice_sample(prop = 0.33) %>%  # 33% from each room type
  ungroup()

var_complete_sample <- hclust(dist(sampled_data[,-10]), method = "complete")
```

#### Elbow method: PCA and cluster vars

```{r}
#| label: within-variation-function
#| echo: true
#| warning: false

within_variation_manual <- function(X, cluster) {
  
  clusters <- unique(cluster)
  results <- data.frame(
    cluster = clusters,
    n = NA,
    within_variation = NA
  )
  
  total_within <- 0
  
  for (k in clusters) {
    
    # Points in cluster k
    Xk <- X[cluster == k, , drop = FALSE]
    
    # Cluster centroid
    centroid <- colMeans(Xk)
    
    # Squared distances to centroid
    squared_distances <- rowSums((sweep(Xk, 2, centroid, "-"))^2)
    
    # Within variation for cluster k
    Wk <- sum(squared_distances)
    
    results[results$cluster == k, "n"] <- nrow(Xk)
    results[results$cluster == k, "within_variation"] <- Wk
    
    total_within <- total_within + Wk
  }
  
  list(
    by_cluster = results,
    total_within_variation = total_within
  )
}
```

```{r}
#| label: wss-complete
#| echo: true
#| warning: false

# complete
WSS_complete_vars <- numeric(9)
for (i in 2:10){
  clusters <- cutree(var_complete_sample, k = i)
  result <- within_variation_manual(scaled_clustering_vars, clusters)
  WSS_complete_vars[i] <- result$total_within_variation
}

WSS_complete_vars <- data.frame(
  k = 2:10,
  WSS = WSS_complete_vars[2:10]
)

ggplot(WSS_complete_vars, aes(x = k, y = WSS)) +
  geom_line(color = "#2c3e50", size = 1) +
  geom_point(color = "#e74c3c", size = 3) +
  scale_x_continuous(breaks = 2:10) +
  labs(
    title = "Elbow Method: Complete Linkage",
    subtitle = "Analysis of Cape Town Airbnb Clusters",
    x = "Number of Clusters (k)",
    y = "Total Within-cluster Variation"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

```

consensus: k = 3

#### Silhouette method

**cluster vars**

```{r}
#complete

fviz_nbclust(sampled_data_scaled, hcut, method = "silhouette", hc_method = "complete")

```

consensus: k =2 - 5

#### Gap statistic

**pca**

```{r}
set.seed(321)
#complete
gap_complete <- clusGap(
  sampled_data_scaled,
  FUNcluster = hcut,
  hc_method = "complete",
  K.max = 10,
  B = 50
)

fviz_gap_stat(gap_complete)

```

consensus: k = 3-4

## Cluster Validation Summary

| Validation Method Complete Linkage |         |
|------------------------------------|:-------:|
| **Elbow Method (WSS)**             | $k = 3$ |
| **Average Silhouette**             | $k = 3$ |
| **Gap Statistic**                  | $k = 3$ |

## Fitting complete Linkage on full dataset and creating 3 clusters

### Cutting the tree

**Clustering vars**

```{r}

cut_vars_tree <- cutree(var_complete_euc, k = 3)
fviz_cluster(list(data = scaled_clustering_vars, cluster = cut_vars_tree),
             geom = "point", 
             ellipse.type = "convex", 
             ggtheme = theme_minimal(),
             main = "Cluster Visualization (Complete Linkage, k=3)")
```

### Silhouette method

**Clustering Vars**

```{r}
sil_obj_vars <- silhouette(cut_vars_tree, dist_matrix)


fviz_silhouette(sil_obj_vars) +
  labs(title = "Silhouette Plot: Complete Linkage (k=3)")


summary(sil_obj_vars)
```

### Cophenetic

**PCA**

**Clustering Vars**

```{r}
cophenetic_cors <- numeric(1)

coph_dist_com <- cophenetic(var_complete_euc)
cophenetic_cors[1]<- cor(dist_matrix, coph_dist_com)

names(cophenetic_cors) <- c("complete")

cophenetic_cors
```

### Cluster Profiling

**Clustering Vars**

```{r}
vars_complete_df <- data.frame(clustering_variables, cluster = cut_vars_tree)

# Calculates the means for every numeric column
vars_complete_means <- aggregate(. ~ cluster, data = vars_complete_df, FUN = mean)

# View the table
print(vars_complete_means)


```

```{r}
library(sf)
library(tmap)
library(dplyr)

map_data <- clustering_var_orig %>%
  select(Longitude, Latitude) %>% 
  mutate(Segment = factor(cut_vars_tree, 
                          levels = c(1, 2, 3), 
                          labels = c("Standard", "Elite", "Budget"))) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

# 2. Reset tmap and plot
tmap_mode("plot")

tm_shape(map_data) +
  tm_dots(col = "Segment", 
          palette = c("blue", "red"),
          size = 0.09,
          alpha = 0.6,
          title = "Airbnb Segments") +
  tm_layout(title = "Cape Town Market Clusters",
            legend.outside = TRUE,
            frame = FALSE)

tmap_mode("view") 


tm_shape(map_data) +
  tm_dots(col = "Segment", 
          palette = c("Standard" = "blue", "Elite" = "deeppink", budget = "orange"),
          size = 0.05, 
          alpha = 0.7,
          title = "Airbnb Segments") + 
  # Coordinates for Cape Town center
  tm_view(set.view = c(18.4233, -33.9249, 11)) 
```

# Non-Hierarchical Clustering

## DBSCAN

### Clustering variables

```{r}
kNNdistplot(scaled_clustering_vars, k = 4)
title(main = "DBSCAN Parameter Selection: k-Distance Plot",
      sub = "Analysis of Cape Town Airbnb Clusters")
#threshold line
abline(h = 0.5, col = "red", lty = 2)
```

```{r}
db <- dbscan(scaled_clustering_vars, eps = 1, MinPts = 5)
db2 <- dbscan(scaled_clustering_vars, eps = 1.5, minPts = 5)
db3 <- dbscan(scaled_clustering_vars, eps = 2.0, minPts = 5)
db4 <- dbscan(scaled_clustering_vars, eps = 2.5, minPts = 5)
db5 <- dbscan(scaled_clustering_vars, eps = 3.0, minPts = 5)
# Compare the counts
table(db$cluster)  # Current
table(db2$cluster) # Result with eps = 1.5
table(db3$cluster) # Result with eps = 2.0
table(db4$cluster) # Result with eps = 2.5
table(db5$cluster) # Result with eps = 3.0

#plot for db5
fviz_cluster(db5, data = clustering_variables[,-10], geom = "point", main = "DBSCAN Cluster plot")
```

```{r}
dist_db5 <- dist(scaled_clustering_vars[db5$cluster != 0, ])
sil_db5 <- silhouette(db5$cluster[db5$cluster != 0], dist_db5)
summary(sil_db5)
```

### Cluster profiling

```{r}
vars_dbscan_df <- data.frame(clustering_variables, cluster = db5$cluster)

# Calculates the means for every numeric column
vars_dbscan_means <- aggregate(. ~ cluster, data = vars_dbscan_df, FUN = mean)

# View the table
print(vars_dbscan_means)
```

#Comparative Analysis of 2 methods \### ARI

```{r}
# Adjusted Rand Index (ARI)
ari_score <- adj.rand.index(cut_vars_tree, db5$cluster)
```

```{r}
#Jaccard
library(fossil)
jaccard_score  <- jaccard(cut_vars_tree, db5$cluster)
```

```{r}
metrics <- cbind(ari_score, jaccard_score)
colnames(metrics) <- c("ARI", "Jaccard Index")
```

# Self Organising Maps

```{r}
#| label: Architecture
#| echo: true
#| warning: false

# How many nodes in the grid? Grid Size
k <- 44
l <- 44
m <- k*l # total number of nodes in grid

# How many weights, each node m_i has p weights
p <- dim(scaled_clustering_vars)[2] # num of variables in input space
w <- p*m

# Create SOM grid
som_grid <- somgrid(xdim = k, ydim=l, topo="hexagonal")
```

```{r}
#| label: som-model
#| echo: true
#| warning: false

set.seed(3145)

model_file <- "som_model.rds"

if (file.exists(model_file)) {
    message("Loading RDS file")
    som_model <- readRDS(model_file)
  } else {
      message("No RDS file found, Computing Model")
      som_model <- som(scaled_clustering_vars, # Input space
                 grid   = som_grid, # Grid space
                 rlen   = 10000, # Number of iterations 
                 alpha  =c(0.05,0.01), # Learning rate, smaller the better, try decreasing over time
                 keep.data = TRUE  )
  
    # Save model as an RDS file
    saveRDS(som_model, model_file)
    }

# Show the average variable values for each of the 400 nodes in the grid
SomModelCodes = som_model$codes[[1]]

# Plot the distributions of each variables from the nodes
for (i in 1:dim(SomModelCodes)[2]){
  plot(SomModelCodes[,i], ylab = colnames(scaled_clustering_vars)[i])
}
```

## Quality Control - For Model

### Training Process

As the model iterates, a node's weights becomes similar to the samples in it. So we expect the distance between node weights and node samples to decrease as the iterations increase until a plateau is reached. If this is not the case, then more iterations are required.

```{r}
#| label: training-changes-plot
#| echo: true
#| warning: false

plot(som_model,type="changes")
```

### Node Counts

Visualization of the number of observations in each node enables map quality assessment. An ideal model has a uniform distribution of observations across the grid. Large values in a few areas indicates a larger map requirement, while a sparse map with empty nodes indicates a smaller grid.

```{r}
#| label: node-counts-plot
#| warning: false
#| include: false

plot(som_model,type="count")
```

Pretty reasonable spread, with only 5% of the nodes containing zero obs

### U-Matrix

The U-Matrix sums up the distances of the observations from all of its immediate neighbours. Low neighbour distance implies very similar observations. Areas of low distance imply that the nodes are very similar to each other.

```{r}
#| label: distance-plot
#| warning: false
#| include: false

plot(som_model, type="dist.neighbours")
```

This is a good distance plot. A lot of the nodes are very similar to their immediate neighbours. Only 3 are dissimilar-reflecting that those observations are naturally very unique.

### Quantization Error

This compares the observations to the vector code representing them. It measures how well the observations are represented by their node (and its vector code).

```{r}
#| label: quantization-plot
#| warning: false
#| include: false

plot(som_model,type="quality")
```

A lot of the nodes have a low distance meaning the observations contained in them are very well represented by their node. This model accurately represents the observations.

## Individual Plots for variables

### Vector Code Plots

As mentioned above, the node vectors are representative of the observations contained in them. They average the observations contained across the variables. The below plot shows the magnitude of each variable in each node e.g the node in the top right has large values for price, revenue, accommodates and are usually entire homes.

```{r}
#| label: vec-code-plot
#| warning: false
#| include: false

plot(som_model, type="codes", palette.name = rainbow)
```

### Heatmaps

Generating heat maps for each variable across the nodes. This shows which nodes contain the highest magnitude for each variable. This is important for investigating the distribution of each variable across the nodes and seeing the profiles of each neighbourhood. Could this be done for two variables at a time?

```{r}
#| label: heatmaps
#| echo: false
#| warning: true

##################################################
# Create Palettes for each of the variables
##################################################
priceRamp <- function(n, alpha = 1) {
  colorRampPalette(
    c(
      "#08306B",  # dark blue
      "#1A9850",  # green
      "#FFFFBF",  # yellow
      "#FDAE61",  # orange
      "#D73027"   # red
    ),
    alpha = FALSE
  )(n)
}

##################################################
# Create heatmaps using the corresponding pallettes
##################################################

par(mfrow = c(3,3), mar = c(1,1,4,1), oma = c(0,0,2,0))
for (i in 1:dim(SomModelCodes)[2]){
  plot(
  som_model,
  type = "property",
  property = SomModelCodes[, i],
  main = "",
  keepMargins = TRUE,
  palette.name = priceRamp
  )
  mtext(
    text = colnames(SomModelCodes)[i],
    side = 3,          # top of plot
    line = 0.5,        # how far above the plot border (increase if still clipped)
    cex  = 0.85,
    font = 2           # bold
  )
}

```

## Creating Clusters from Self Organising Map

```{r}
#| label: clusters-from-som
#| echo: false
#| warning: false

# # of observations to cluster
length(som_model$unit.classif)

# Show assignment of last 20 observations
tail(som_model$unit.classif, 20)

# Show table of nodes and number of observations in them
node_counts <- table(som_model$unit.classif)
mean(node_counts)

# Compute summary of quantization error
summary(
  sqrt(
    rowSums(
      (scaled_clustering_vars - som_model$codes[[1]][som_model$unit.classif, ])^2
    )
  )
)
```

### Apply K-means Clustering to Code vectors:

Applying k-means determines the number of clusters to specify the nodes. Split the nodes into K clusters and determine the within sum of squares to maximize how similar the nodes within each cluster are to each other. From the plot, we see a reasonable K to select is 3 clusters.

```{r}
#| label: code-vector-kmeans
#| echo: false
#| warning: false

# Compute within sum of squares for different values of K-clusters
set.seed(3145)
wss <- (nrow(SomModelCodes)-1)*sum(apply(SomModelCodes,2,var)) 
for (i in 2:15) {
  wss[i] <- sum(kmeans(SomModelCodes, centers=i)$withinss)
}
plot(wss, type = "b")
```

### Apply (Complete) Hierarchical Clustering To Code Vectors

```{r}
#| label: HC-on-code-vectors
#| echo: false
#| warning: false

SomModelCodes = som_model$codes[[1]]
# Compute distances between SOM codebook vectors
som_dist <- dist(SomModelCodes, method = "euclidean") 

set.seed(3145)
# Perform hierarchical clustering on those distances
hierarchicalcluster <- hclust(som_dist, method = "complete") 

# Add a cluster column to the vector codes
cluster <- c(0)
SomModelCodes <- cbind(SomModelCodes, cluster)

# different number of clusters:
totwithinss = NA
wss0 <- sum(rowSums((sweep(SomModelCodes, 2, colMeans(SomModelCodes), "-"))^2))
totwithinss[1] = wss0


for (k in 2:15){
  cluster = cutree(hierarchicalcluster, k=k) # obtain new clusters 
  SomModelCodes[,10] <- cluster # assign new clusters to code vectors
  
  # group by cluster k, compute ss for that cluster, add to tss
  kss <- 0
  for (c in 1:k){ # group each cluster and compute its ss
    current_cluster <- SomModelCodes[SomModelCodes[, dim(SomModelCodes)[2]] == c, 1:6,drop = FALSE]
    kss <- kss + rowSums((sweep(current_cluster, 2, colMeans(current_cluster), "-"))^2)
  }
  totwithinss[k] <- kss # Add this k's ss to the vector of ss's
}

# Create a dataframe of k's and their corresponding within SS
wss_df <- data.frame(
  k = 2:15,
  WSS = totwithinss[-1]
)

# Plot the K and its within SS
ggplot(wss_df, aes(x = k, y = WSS)) +
  geom_line() +
  geom_point(size = 3) +
  scale_x_continuous(breaks = 1:15) +
  labs(
    title = "Elbow Plot for Complete Linkage Clustering",
    x = "Number of clusters (K)",
    y = "Total within-cluster sum of squares"
  ) +
  theme_minimal()
```

## NbClust

```{r}
#| label: som-nbclust
#| echo: false

SomModelCodes = som_model$codes[[1]]
result <- NbClust(data = SomModelCodes, 
                  distance = "euclidean",
                  min.nc = 2, max.nc = 10,
                  method = "complete")
```

## Silhouette Plots

```{r}
#| label: som-sil
#| echo: false
#| warning: false

SomModelCodes = som_model$codes[[1]]
fviz_nbclust(SomModelCodes, hcut, method = "silhouette", hc_method = "complete")
```

## Gap Analysis

```{r}
#| label: som-gap
#| echo: false
#| warning: false

SomModelCodes = som_model$codes[[1]]
set.seed(321)
#complete
gap_som <- clusGap(
  SomModelCodes,
  FUNcluster = hcut,
  hc_method = "complete",
  K.max = 10,
  B = 50
)

fviz_gap_stat(gap_som)
```

### Add clusters from HC back to SOM

```{r}
#| label: plot-som-with-clusters
#| warning: true
#| include: false

# get clusters for each node
som.hc <- cutree(hierarchicalcluster, k = 4)

# get cluster assignments for each obs
cluster_assignments <- som.hc[som_model$unit.classif]

# Add clusters back to original variables
# convert data to df
clustering_vars_df <- as.data.frame(clustering_variables)

# Assign cluster column
clustering_vars_df$cluster <- as.factor(cluster_assignments)

```

```{r}
#| label: viz-som-clusters
#| echo: false
#| warning: false

fviz_cluster(
  list(
    data    = clustering_vars_df[, !names(clustering_vars_df) %in% "cluster"],
    cluster = cluster_assignments
  ),
  geom         = "point",
  ellipse.type = "convex",
  main         = "Hierarchical Clusters on SOM (PCA projection)"
)
```

```{r}
#| label: cluster profiles
#| echo: false
#| warning: false

cluster_profiles <- clustering_vars_df %>%
  group_by(cluster) %>%
  summarise(
    across(where(is.numeric), mean, .names = "{.col}"),
    n = n()
  ) %>%
  mutate(pct = round(n / sum(n) * 100, 1))

print(cluster_profiles)
```

#Comparative Analysis of 3 methods \### ARI

```{r}
#| label: som vs HC
#| echo: false
# Adjusted Rand Index (ARI) between HC & SOM 
ari_HC_SOM <- adj.rand.index(cut_vars_tree, cluster_assignments)
ari_HC_SOM

# Jaccards between HC & SOm
jaccard_HC_SOM  <- jaccard(cut_vars_tree, cluster_assignments)
jaccard_HC_SOM
```

```{r}
#| label: som vs DBSCAN
#| echo: false

# Adjusted Rand Index (ARI) between HC & SOM 
ari_DBSCAN_SOM <- adj.rand.index(db5$cluster, cluster_assignments)
ari_DBSCAN_SOM

# Jaccards between HC & SOm
jaccard_DBSCAN_SOM  <- jaccard(db5$cluster, cluster_assignments)
jaccard_DBSCAN_SOM
```

```{r}
metrics <- cbind(ari_score, jaccard_score)
colnames(metrics) <- c("ARI", "Jaccard Index")
```

+------------------------------------------+---------------+
| Validation Method - Self Organising Maps |               |
+------------------------------------------+---------------+
| **Elbow Method (WSS)**                   | $k=2$         |
|                                          |               |
| -   K-means                              | $k=9$         |
|                                          |               |
| -   Hierarchical (Complete)              |               |
+------------------------------------------+---------------+
| **Average Silhouette**                   | $k=4$         |
+------------------------------------------+---------------+
| **Gap Statistic**                        | $k= 3$        |
+------------------------------------------+---------------+
